---
title: "R Data Structures"
output: md_document
---

# R Data Structure

This section will go over some R data structures and how to work with them.

## Types

R utilizes several data types. While R is pretty relaxed with types, this can cause some issues, so it is important to be aware of them. However, I will only cover some of the basics here (and in a simplified manner).

### Character

Characters are both singular characters or strings of characters, such as

```{r}
a_letter <- "A"
a_string <- "A string"

typeof(a_letter)
typeof(a_string)
```

Lists which have elements of a certain type inherit that type

```{r}
list_of_strings <- c("A", "list", "of", "strings")

typeof(list_of_strings)
```

#### Factors

An interesting spin-off of the character or string is a `factor`. Factors are used to represent the various levels of a categorical variable. Because of this, factor *levels* are unique (not repeated).

```{r}
some_factors <- as.factor(c('bacteria', 'bacteria', 'virus', 'fungus', 'virus'))
                          
some_factors
```


### Numbers

When storing a numerical value, its type defaults to a double. Numbers without a decimal value can also be stored as an integer.

```{r}
a_number <- 1.5

typeof(a_number)

another_number <- 1

typeof(another_number)

an_integer <- integer(1)

typeof(an_integer)
```

### Logical

Boolean values are either TRUE or FALSE. These are integral to logic statements, which is why they are the *logical* type in R.

```{r}
a_true <- TRUE

typeof(a_true)

a_false <- FALSE

typeof(a_false)
```

They can also be represented using the shorthand T or F

```{r}
a_true <- T

typeof(a_true)

a_false <- F

typeof(a_false)
```


## Structures

R would be useless if there weren't powerful ways of organizing and storing data.

### Lists

The most basic structure is the list. Lists can store elements of any data type.

```{r}
string_list <- list("Legolas", "Gimli", "Sauron")
str(string_list) # str can be used to get a more readable summary
typeof(string_list)

number_list <- list(1, 6, 12)
str(number_list)
typeof(number_list)

bool_list <- list(F, T, F)
str(bool_list)
typeof(bool_list)
```
You can even have a list of lists.

```{r}
list_list <- list(list(1, 2), list("a", "b"))

str(list_list)

typeof(list_list)
```

#### Vectors

A special type of list is a vector, which can be defined with the `c()` shorthand.

Vectors inherit the type of their elements.

```{r}
number_vector <- c(2, 1, 4)
typeof(number_vector)

string_vector <- c("foo", "bar", "baz", "quux")
typeof(string_vector)

bool_vector <- c(T, F, F)
typeof(bool_vector)
```

Lists can have elements of different types. Vectors will coerce elements to a consistent type.

```{r}
coerced_list <- c(1, F, "foo")

str(coerced_list)

typeof(coerced_list)

mixed_list <- list(1, F, "foo")

str(mixed_list)

typeof(mixed_list)
```

### Dataframe

Lists are great, but for data analysis, we need something bigger and better. This is the dataframe. A dataframe can be thought of kind of as a table or a spreadsheet.

```{r}
a_df <- data.frame(plant = c("pingicula", "drosera",
                         "dionaea", "drosera", "dionaea"),
                   length = c(66, 82, 73, 49, 65),
                   cool = c(F, T, T, T, T)
                   )

a_df
```

## Accessing Data

### Lists

List and vector elements are accessed using square brackets `[]` In the square bracket, you place the *index* to retrieve. Unlike in most programming languages, indices start at 1 instead of 0, which would be convenient if it didn't feel so awkward coming from other languages!

```{r}
string_vector

string_vector[1]
```
List slices can also be accessed by using the notation `[start:end]`. Again, unlike in many other languages, the range is inclusive (the value at the end index is returned).

```{r}
string_vector[1:2] # First through second elements
```
In order to exclude instead of retrieve certain elements of a list, you can place a negative index value.

```{r}
string_vector[-3] # All elements but the third
```
Additionally, a vector of boolean values can be used to retrieve certain elements. In this case, it is best that the boolean vector has the same length as the list.

Indices where there is True value are returned, indices with a False value are not.

```{r}
string_vector[c(T, F, F, T)]
```

When using boolean vectors that are not the same length as the list, you have to be careful because the list is "recycled" without any warning.

```{r}
string_vector[c(T, F)] # Returns every other list element
```

The fact that you can access list elements with boolean vectors is really useful. For instance, you can build a vector based on a conditional, then use that to subset the list.

```{r}
# Create vector of booleans
# The value will be True when the string_vector element is not "bar
string_vector != "bar"


# I read the following as string_vector where string_vector is not equal to "bar"
string_vector[string_vector != "bar"]
```
### Dataframes

Data frame parts can also be accessed using `[]` notation, but row and column indices are specified as `[row, column]`. All list subsetting rules described above apply to rows and columns, because each row and column can be thought of as a list.

```{r}
a_df

# Access a single value
a_df[4, 1] # Row 4, column 1

# Access a whole row
a_df[4, ]

# Access a whole column
a_df[ , 3]
```

Columns can also be accessed using strings as indices. A certain type of list, *named list*, can also be accessed this way.

```{r}
a_df["length"]
```

A simplified notation for this is also available, but note that the returned type is a vector.

```{r}
a_df$length

typeof(a_df["length"])
typeof(a_df$length)
```

